---
title: "Create UniProt entry mapping between RefSeq 20160914 and UniProt 2021_02"
output:
    html_document:
        df_print: paged
---

```{r, message=FALSE}
library(here)
library(tidyverse)
```

Source here is RefSeq 20160914 and Target is UniProt 2021_02
```{r}
source_annotation_tbl = read_tsv(here('annotation/refseq_20160914/tracked_results/refseq_20160914_human_only_unique_loci.tsv.gz'))
target_tbl = read_tsv(here('annotation/mapping_uniprot_2021_02/tracked_results/uniprot_reviewed_human_proteome_hgnc_only.v2021_02.tsv.gz'))
```

```{r}
colnames(source_annotation_tbl)
colnames(target_tbl)
```

Some source protein entries lack HGNC ID
```{r}
source_annotation_tbl %>%
    filter(is.na(hgnc_id)) %>%
    nrow()
```

Find the matching records that has identical protein sequence in the target set
```{r}
mapping_tbl.identical = source_annotation_tbl %>%
    select(refseq_prot_id, hgnc_id, source_uniparc_id = uniparc_id) %>%
    inner_join(target_tbl %>% distinct(uniprot_acc, target_uniparc_id = uniparc_id), 
               by = c('source_uniparc_id' = 'target_uniparc_id')) %>%
    mutate(
        target_uniparc_id = source_uniparc_id,
        mapping_approach = 'identical_sequence'
    )

mapping_tbl.identical %>% head()
```
```{r}
source_annotation_tbl.remaining = source_annotation_tbl
source_annotation_tbl.remaining = source_annotation_tbl.remaining %>%
    filter(!refseq_prot_id %in% mapping_tbl.identical$refseq_prot_id)
```


The remaining entries will require sequence alignment (choosing the ones with the same HGNC ID)
```{r}
mapping_tbl.align = source_annotation_tbl.remaining %>%
    select(refseq_prot_id, hgnc_id, source_uniparc_id = uniparc_id) %>%
    inner_join(
        target_tbl %>% 
            select(uniprot_acc, hgnc_id = hgnc_gene_id, target_uniparc_id = uniparc_id),
        by = "hgnc_id"
    ) %>%
    mutate(mapping_approach = 'global_seq_align')

mapping_tbl.align %>% head()
```
```{r}
source_annotation_tbl.remaining = source_annotation_tbl.remaining %>%
    filter(!refseq_prot_id %in% mapping_tbl.align$refseq_prot_id)

source_annotation_tbl.remaining %>%
    count(!is.na(hgnc_id))
```

```{r}
mapping_tbl.align_by_symbol = source_annotation_tbl.remaining %>%
    select(refseq_prot_id, hgnc_id, symbol, source_uniparc_id = uniparc_id) %>%
    inner_join(
        target_tbl %>% 
            select(uniprot_acc, symbol = hgnc_gene_name, target_uniparc_id = uniparc_id),
        by = "symbol"
    ) %>%
    mutate(mapping_approach = 'global_seq_align')

mapping_tbl.align_by_symbol %>% head()
```
```{r}
source_annotation_tbl.failed = source_annotation_tbl.remaining %>%
    filter(!refseq_prot_id %in% mapping_tbl.align_by_symbol$refseq_prot_id)
```

```{r}
source_annotation_tbl.failed %>%
    count(!is.na(hgnc_id))
```

Make the mapping table
```{r}
mapping_tbl = bind_rows(
    mapping_tbl.identical,
    mapping_tbl.align,
    mapping_tbl.align_by_symbol %>% select(-symbol)
) %>%
    arrange(refseq_prot_id)

mapping_tbl %>% head()
```

Some source proteins can be mapped to multiple targets:
```{r}
num_source_ids_with_dup_mapping = mapping_tbl %>%
    count(refseq_prot_id, sort = TRUE) %>%
    filter(n > 1)

num_source_ids_with_dup_mapping %>% head()
```

Showing some statistics
```{r, echo=FALSE, collapse=TRUE}
cat("Total source entries:", nrow(source_annotation_tbl), "\n")
cat("Total target entries:", nrow(target_tbl), "\n")

cat("Source entries without any mapping:", nrow(source_annotation_tbl.failed), "\n")
cat("Mapping table (unique source ID) entry pairs:", n_distinct(mapping_tbl$refseq_prot_id), "\n")
cat("Mapping type breakdown (unique source ID):\n")
mapping_tbl %>%
    distinct(refseq_prot_id, .keep_all = TRUE) %>%
    count(mapping_approach) %>%
    format() %>%
    cat(sep = "\n")

cat("Mapping table (with duplication) entry pairs:", nrow(mapping_tbl), "\n")
cat("    ... source entries with duplicated mapping:", nrow(num_source_ids_with_dup_mapping), "\n")
mapping_tbl %>%
    count(mapping_approach) %>%
    format() %>%
    cat(sep = "\n")
```

```{r}
mapping_tbl %>% 
    write_tsv('../tracked_results/mappings/refseq_20160914_to_uniprot_2021_02_mappable.tsv.gz')

source_annotation_tbl.failed %>%
    write_tsv('../tracked_results/mappings/refseq_20160914_to_uniprot_2021_02_not_mappable.tsv.gz')
```

