---
title: "R Notebook"
output:
    html_document:
        df_print: paged
---

```{r, message=FALSE}
library(SummarizedExperiment)
library(cmapR)
library(here)
library(tidyverse)
```

```{r}
DATA_ROOT = '~/Box/MyCPTAC/CPTAC_proteome_v2.1'
```

Load extra functions for mapping
```{r}
source('utils.R')
```

```{r}
protein_pth = fs::path(DATA_ROOT, 'CPTAC2/BRCA_prospective/prosp-brca-v5.3-proteome-ratio-norm-NArm.gct')

phospho_pth = fs::path(DATA_ROOT, 'CPTAC2/BRCA_prospective/prosp-brca-v5.3-phosphoproteome-ratio-norm-unfiltered.gct')
phospho_loc_pth = fs::path(DATA_ROOT, 'CPTAC2/BRCA_prospective/prosp-brca-v5.3-phosphoproteome-ratio-norm-unfiltered.peptide_location.tsv.gz')

acetyl_pth = fs::path(DATA_ROOT, 'CPTAC2/BRCA_prospective/prosp-brca-v5.3-acetylome-ratio-norm-unfiltered.gct')
acetyl_loc_pth = fs::path(DATA_ROOT, 'CPTAC2/BRCA_prospective/prosp-brca-v5.3-acetylome-ratio-norm-unfiltered.peptide_location.tsv.gz')

se_metadata = list(
    cohort = 'CPTAC2 BRCA prospective cohort',
    pipeline = 'Broad',
    mapped_uniprot_release = '2021_02'
)
```

```{r}
protein_annotation_gr = read_rds(here(
    'annotation/refseq_20160914/tracked_results/refseq_20160914_human_only_unique_loci_GRanges.rds'
))
names(mcols(protein_annotation_gr))
```

Read UniProt coord_mapping
```{r}
uniprot_coord_mapping_tbl = read_refseq_to_uniprot_coord_mapping(
    refseq_ver = "20160914",
    uniprot_ver = se_metadata$mapped_uniprot_release
)
```


## Protein
```{r}
protein_gct = parse_gctx(protein_pth)
raw_protein_se = as(protein_gct, "SummarizedExperiment")
```

Make sure the gene symbols are the same as the protein annotation
```{r}
rowData(raw_protein_se) %>%
    as_tibble() %>%
    select(accession_number, geneSymbol) %>%
    inner_join(protein_annotation_gr %>% mcols() %>% as_tibble(),
              by = c('accession_number' = 'refseq_prot_id')) %>%
    filter(geneSymbol != symbol)
```

Some proteins are not of RefSeq
```{r}
prot_ids_not_in_refseq = setdiff(
    rowData(raw_protein_se)$accession_number,
    protein_annotation_gr$refseq_prot_id
)

rowData(raw_protein_se)[prot_ids_not_in_refseq, ] %>%
    as_tibble()
```

```{r}
protein_row_tbl = rowData(raw_protein_se) %>%
    as_tibble() %>%
    # Keep only RefSeq proteins
    filter(accession_number %in% protein_annotation_gr$refseq_prot_id) %>%
    mutate(subgroupNum = as.character(subgroupNum)) %>%
    extract(
        subgroupNum,
        c('protein_grp', 'protein_subgrp'),
        "^(\\d+)\\.(\\d+)",
        convert = TRUE, remove = FALSE
    )
```

By default genes are not unique (isoform level)
```{r}
protein_row_tbl %>% count(geneSymbol, sort = TRUE) %>%
    filter(n > 1)
```

Convert to gene level
```{r}
sgt_protein_row_tbl = protein_row_tbl %>%
    arrange(geneSymbol, protein_subgrp) %>%
    group_by(geneSymbol) %>%
    filter(row_number() == 1)

sgt_protein_row_tbl %>% head()
```
Add gene level annotation
```{r}
gene_annotation_tbl = protein_annotation_gr %>%
    as_tibble() %>%
    distinct(symbol, hgnc_id, entrez_gene_id, ensembl_gene_id)

sgt_protein_full_row_tbl = sgt_protein_row_tbl %>%
    select(
        symbol = geneSymbol,
        refseq_prot_id = accession_number,
        numColumnsProteinObserved, numSpectraProteinObserved,
        protein_mw, percentCoverage,
        numPepsUnique, scoreUnique, subgroupNum
    ) %>%
    left_join(gene_annotation_tbl, by = "symbol")

sgt_protein_full_row_tbl %>% head()
```
Organize colData
```{r}
protein_col_tbl = colData(raw_protein_se) %>%
    as_tibble(rownames = 'preferred_sample_name') %>%
    select(
        preferred_sample_name, case_id = Participant,
        aliquot_id = Human.Readable.Label,
        tmt_experiment = Experiment, tmt_channel = Channel,
        sample_type = Type,
        tumor_id = Tumor.ID, normal_id = Normal.ID,
        rnaseq_id = RNASeq.ID
    ) %>%
    mutate(
        case_id = str_sub(case_id, start = 2L),
        sample_type = case_when(
            sample_type == 'Tumor' ~ 'Primary Tumor',
            TRUE ~ NA_character_
        )
    )

protein_col_tbl %>% head()
```
```{r}
protein_se = raw_protein_se[sgt_protein_full_row_tbl$refseq_prot_id, ]
rownames(protein_se) = sgt_protein_full_row_tbl$symbol
rowData(protein_se) = sgt_protein_full_row_tbl %>% column_to_rownames('symbol')
colData(protein_se) = protein_col_tbl %>% column_to_rownames('preferred_sample_name') %>% as('DataFrame')
assayNames(protein_se) = 'normalized'

metadata(protein_se) = se_metadata
```

```{r}
protein_se
```



## Phospho
```{r}
phospho_gct = parse_gctx(phospho_pth)
raw_phospho_se = as(phospho_gct, "SummarizedExperiment")

phospho_loc_tbl = read_tsv(
    phospho_loc_pth,
    col_types = cols(
        peptide_start = col_integer(),
        peptide_end = col_integer()
    ))
```

```{r}
raw_phospho_row_tbl = rowData(raw_phospho_se) %>%
    as_tibble(rownames = 'original_id') %>%
    inner_join(phospho_loc_tbl, by = c('original_id' = 'rid'))

raw_phospho_row_tbl %>% head()
```
```{r}
phospho_na_percent = rowSums(is.na(assay(raw_phospho_se[raw_phospho_row_tbl$original_id, ]))) / ncol(raw_phospho_se)
hist(phospho_na_percent)
```

```{r}
phospho_row_tbl = raw_phospho_row_tbl %>%
    bind_cols(tibble(na_percent = phospho_na_percent)) %>%
    filter(
        peptide_loc_valid,
        # na_percent <= 0.8
    ) %>%
    select(
        original_id,
        refseq_prot_id = accession_number,
        phosphosites,
        symbol = geneSymbol,
        peptide = sequence,
        peptide_start,
        peptide_end,
        na_percent
    ) %>%
    arrange(symbol, phosphosites, na_percent) %>%
    mutate(
        unique_site_id = base::make.unique(
            str_c(symbol, refseq_prot_id, phosphosites, sep = ':'),
            sep='.'
        )
    )

phospho_row_tbl %>% head()
```
Some sites have duplication
```{r}
phospho_row_tbl %>%
    count(symbol, phosphosites, sort = TRUE) %>%
    filter(n > 1)
```

Combine the RefSeq annotation
```{r}
protein_annotation_tbl = protein_annotation_gr %>%
    mcols() %>%
    as_tibble() %>%
    select(-symbol)
```

```{r}
phospho_row_full_tbl = phospho_row_tbl %>%
    select(unique_site_id, original_id, refseq_prot_id, symbol, phosphosites,
           peptide, peptide_start, peptide_end) %>%
    left_join(
        protein_annotation_tbl,
        by = 'refseq_prot_id'
    ) %>%
    annotate_uniprot_mapping("phosphosites", uniprot_coord_mapping_tbl)

phospho_row_full_tbl %>% head()
```

```{r}
identical(colnames(raw_phospho_se), protein_col_tbl$preferred_sample_name)
setdiff(colnames(raw_phospho_se), protein_col_tbl$preferred_sample_name)
```

raw tables contained adjacent normals. excluded the normals and take the first replicate.

```{r}
phospho_se = raw_phospho_se[phospho_row_full_tbl$original_id, protein_col_tbl$preferred_sample_name]

rownames(phospho_se) = phospho_row_full_tbl$unique_site_id
rowData(phospho_se) = phospho_row_full_tbl %>% column_to_rownames('unique_site_id')
colData(phospho_se) = colData(protein_se)
assayNames(phospho_se) = 'normalized'

metadata(phospho_se) = se_metadata
```

```{r}
phospho_se
```



## Acetyl
```{r}
acetyl_gct = parse_gctx(acetyl_pth)
raw_acetyl_se = as(acetyl_gct, "SummarizedExperiment")

acetyl_loc_tbl = read_tsv(
    acetyl_loc_pth,
    col_types = cols(
        peptide_start = col_integer(),
        peptide_end = col_integer()
    ))
```

```{r}
raw_acetyl_row_tbl = rowData(raw_acetyl_se) %>%
    as_tibble(rownames = 'original_id') %>%
    inner_join(acetyl_loc_tbl, by = c('original_id' = 'rid'))

raw_acetyl_row_tbl %>% head()
```
```{r}
acetyl_na_percent = rowSums(is.na(assay(raw_acetyl_se[raw_acetyl_row_tbl$original_id, ]))) / ncol(raw_acetyl_se)
hist(acetyl_na_percent)
```
```{r}
acetyl_row_tbl = raw_acetyl_row_tbl %>%
    bind_cols(tibble(na_percent = acetyl_na_percent)) %>%
    filter(
        peptide_loc_valid,
        # na_percent <= 0.8
    ) %>%
    select(
        original_id,
        refseq_prot_id = accession_number,
        acetylsites,
        symbol = geneSymbol,
        peptide = sequence,
        peptide_start,
        peptide_end,
        na_percent
    ) %>%
    arrange(symbol, acetylsites, na_percent) %>%
    mutate(
        unique_site_id = base::make.unique(
            str_c(symbol, refseq_prot_id, acetylsites, sep = ':'),
            sep='.'
        )
    )

acetyl_row_tbl %>% head()
```

Some sites have duplication
```{r}
acetyl_row_tbl %>%
    count(symbol, acetylsites, sort = TRUE) %>%
    filter(n > 1)
```


Combine the RefSeq annotation
```{r}
acetyl_row_full_tbl = acetyl_row_tbl %>%
    select(unique_site_id, original_id, refseq_prot_id, symbol, acetylsites,
           peptide, peptide_start, peptide_end) %>%
    left_join(
        protein_annotation_tbl,
        by = 'refseq_prot_id'
    ) %>%
    annotate_uniprot_mapping("acetylsites", uniprot_coord_mapping_tbl)

acetyl_row_full_tbl %>% head()
```
Check if we can re-use the protein col data
```{r}
identical(colnames(raw_acetyl_se), protein_col_tbl$preferred_sample_name)
setdiff(colnames(raw_acetyl_se), protein_col_tbl$preferred_sample_name)
```

```{r}
acetyl_se = raw_acetyl_se[acetyl_row_full_tbl$original_id, protein_col_tbl$preferred_sample_name]

rownames(acetyl_se) = acetyl_row_full_tbl$unique_site_id
rowData(acetyl_se) = acetyl_row_full_tbl %>% column_to_rownames('unique_site_id')
colData(acetyl_se) = colData(protein_se)
assayNames(acetyl_se) = 'normalized'

metadata(acetyl_se) = se_metadata
```

```{r}
acetyl_se
```



## Export
```{r}
fs::dir_create(here('final_output/CPTAC2_BRCA_prospective'))
protein_se %>% saveRDS(here('final_output/CPTAC2_BRCA_prospective/CPTAC2_BRCA_prospective_broad_proteome.v2.0.rds'))
phospho_se %>% saveRDS(here('final_output/CPTAC2_BRCA_prospective/CPTAC2_BRCA_prospective_broad_phosphoproteome.v2.2.rds'))
acetyl_se %>% saveRDS(here('final_output/CPTAC2_BRCA_prospective/CPTAC2_BRCA_prospective_broad_acetylome.v2.2.rds'))
```
