---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---

```{r, message=FALSE}
library(SummarizedExperiment)
library(here)
library(readxl)
library(tidyverse)
```

```{r}
DATA_ROOT = '~/Box/MyCPTAC/CPTAC_data_collection_v1'
```

```{r}
tmt_pth = fs::path(DATA_ROOT, 'CPTAC2/CRC_prospective/TMT10_Label_to_Sample_Mapping_File_COprospective_Protoeme_PNNL_S037.xlsx')
sample_type_pth = fs::path(DATA_ROOT, 'CPTAC2/CRC_prospective/pdc_graphql_crc_specimen.tsv.gz')

protein_pth = fs::path(DATA_ROOT, 'CPTAC2/CRC_prospective/CPTAC2_Colon_Prospective_Collection_PNNL_Proteome.tmt10.tsv.gz')

phospho_pth = fs::path(DATA_ROOT, 'CPTAC2/CRC_prospective/CPTAC2_Colon_Prospective_Collection_PNNL_Phosphoproteome.phosphosite.tmt10.tsv.gz')
phospho_loc_pth = fs::path(DATA_ROOT, 'CPTAC2/CRC_prospective/CPTAC2_Colon_Prospective_Collection_PNNL_Phosphoproteome.phosphosite.tmt10.tsv.peptide_location.tsv.gz')

se_metadata = list(
    cohort = 'CPTAC2 CRC prospective cohort',
    pipeline = 'CDAP'
)
```


## TMT map
```{r}
tmt_map_tbl = read_excel(tmt_pth) %>%
    filter(PCC == 'PNNL') %>%
    select(folder = `Folder Name`, starts_with('TMT10')) %>%
    pivot_longer(-folder, names_to = c('tmt_channel', 'id_type'), names_pattern = '^TMT10-(\\d+[NC]?) (.*)$') %>%
    pivot_wider(names_from = 'id_type', values_from = 'value') %>%
    filter(!`Participant ID` %in% c('ColonRef')) %>%
    mutate(tmt_experiment = as.integer(str_match(folder, '^(\\d+)CPTAC_')[, 2]))
    
tmt_map_tbl %>% head()
```

```{r}
sample_type_tbl = read_tsv(sample_type_pth)
sample_type_tbl %>% head()
```
```{r}
tmt_map_clean_tbl = tmt_map_tbl %>%
    left_join(sample_type_tbl, 
              by = c("Participant ID"="case_submitter_id", 
                     "Specimen Label"="aliquot_submitter_id")) %>%
    select(case_id = `Participant ID`, aliquot_id = `Specimen Label`,
           tmt_experiment, tmt_channel, sample_type) %>%
    mutate(preferred_sample_name = case_when(
        sample_type == 'Primary Tumor' ~ str_c(case_id, '-T'),
        sample_type == 'Solid Tissue Normal' ~ str_c(case_id, '-N'),
        TRUE ~ NA_character_
    ))

tmt_map_clean_tbl %>% head()
```
```{r}
protein_annotation_gr = read_rds(here(
    'annotation/refseq_20180629/tracked_results/refseq_20180629_human_only_unique_loci_GRanges.rds'
))
names(mcols(protein_annotation_gr))
```

```{r}
gene_annotation_tbl = protein_annotation_gr %>%
    as_tibble() %>%
    distinct(symbol, hgnc_id, entrez_gene_id, ensembl_gene_id)
```


## Protein
```{r}
protein_tbl = read_tsv(protein_pth)
protein_tbl %>% head()
```


```{r}
protein_mat = protein_tbl %>% 
    select(ends_with('Unshared Log Ratio')) %>%
    filter(row_number() > 3) %>%
    rename_all(~ str_sub(.x, end = -str_length('Unshared Log Ratio') - 2)) %>%
    as.matrix()

rownames(protein_mat) = as.character(protein_tbl$NCBIGeneID[-1:-3])

protein_mat[1:5, 1:5]
```

```{r}
protein_row_tbl = protein_tbl %>%
    filter(row_number() > 3) %>%
    select(entrez_gene_id = NCBIGeneID, incompat_symbol_in_manuscript = Gene) %>%
    mutate(entrez_gene_id = as.character(entrez_gene_id)) %>%
    left_join(
        gene_annotation_tbl,
        by = 'entrez_gene_id'
    )
```

Some genes are not covered in RefSeq
```{r}
protein_row_tbl %>%
  filter(is.na(symbol))
```

Genes are unique:
```{r}
protein_row_tbl %>%
    count(symbol) %>%
    filter(n > 1)
```

Some symbol doesn't match the database
```{r}
protein_row_tbl %>%
    filter(incompat_symbol_in_manuscript != symbol)
```

Only select the genes that are in the database
```{r}
selected_protein_row_tbl = protein_row_tbl %>% 
    filter(!is.na(symbol)) %>%
    select(symbol, hgnc_id, entrez_gene_id, ensembl_gene_id, 
           incompat_symbol_in_manuscript)

selected_protein_row_tbl %>% head()
```

Make sure all the samples are in the TMT table
```{r}
setdiff(colnames(protein_mat), tmt_map_clean_tbl$aliquot_id)
```

```{r}
# Re-order the samples and subset the peptide
selected_protein_mat = protein_mat[selected_protein_row_tbl$entrez_gene_id, tmt_map_clean_tbl$aliquot_id]
rownames(selected_protein_mat) = selected_protein_row_tbl$symbol
colnames(selected_protein_mat) = tmt_map_clean_tbl$preferred_sample_name

selected_protein_mat[1:5, 1:5]
```

```{r}
protein_sample_tbl = tmt_map_clean_tbl %>% 
    select(
        preferred_sample_name, case_id, sample_type,
        aliquot_id, tmt_experiment, tmt_channel
    )

protein_sample_tbl %>% head()
```

```{r}
protein_se = SummarizedExperiment(
    assays = list('normalized' = selected_protein_mat),
    rowData = selected_protein_row_tbl %>% column_to_rownames('symbol'),
    colData = protein_sample_tbl %>% column_to_rownames('preferred_sample_name'),
    metadata = se_metadata
)

protein_se
```



### Phospho
```{r}
raw_phospho_tbl = read_tsv(phospho_pth) %>%
    mutate(original_id = str_c(Peptide, Phosphosite, sep = ':'))
```
```{r}
raw_phospho_tbl %>% count(Peptide) %>% filter(n > 1) 
```

Read peptide location
```{r}
phospho_loc_tbl = read_tsv(
    phospho_loc_pth,
    col_types = cols(
        peptide_start = col_integer(), 
        peptide_end = col_integer()
    ))
phospho_loc_tbl %>% head()
```
Some entries are discarded because they are not in RefSeq 2018
```{r}
raw_phospho_tbl %>%
    filter(!original_id %in% phospho_loc_tbl$original_id) %>%
    extract(Phosphosite, c('refseq_prot_id', 'phosphosites'), '^([^:]+):((?:[a-z\\d+])+)$') %>%
    count(Gene, refseq_prot_id, sort = TRUE)
```

```{r}
raw_phospho_row_tbl = raw_phospho_tbl %>%
    transmute(original_id = str_c(Peptide, Phosphosite, sep = ':')) %>%
    inner_join(phospho_loc_tbl, by = 'original_id')
```


Create the phospho matrix
```{r}
phospho_mat = raw_phospho_tbl %>% 
    select(ends_with('Log Ratio')) %>%
    rename_all(~ str_sub(.x, end = -str_length('Log Ratio') - 2)) %>%
    as.matrix()

# Re-order the samples
phospho_mat = phospho_mat[, protein_sample_tbl$aliquot_id]
colnames(phospho_mat) = protein_sample_tbl$preferred_sample_name
rownames(phospho_mat) = raw_phospho_tbl$original_id

phospho_mat = phospho_mat[raw_phospho_row_tbl$original_id, ]
phospho_mat[1:5, 1:10]
```

Calculate the NA% per peptide (to order the site)
```{r}
phospho_na_percent = rowSums(is.na(phospho_mat)) / ncol(phospho_mat)
hist(phospho_na_percent)
```

Construct the phospho rowData table
```{r}
raw_phospho_annotate_row_tbl = raw_phospho_row_tbl %>%
    bind_cols(tibble(na_percent = phospho_na_percent)) %>%
    # Report unique peptide
    group_by(peptide) %>%
    filter(row_number() == 1) %>%
    ungroup() %>%
    filter(
        peptide_loc_valid,
        na_percent <= 0.8
    )

raw_phospho_annotate_row_tbl %>% head()
```

Combine the RefSeq annotation
```{r}
protein_annotation_tbl = protein_annotation_gr %>%
    mcols() %>%
    as_tibble()
```


```{r}
phospho_row_full_tbl = raw_phospho_annotate_row_tbl %>%
    left_join(
        protein_annotation_tbl,
        by = 'refseq_prot_id'
    ) %>%
    # Order the phospho site based on the NA%
    arrange(symbol, phosphosites, na_percent) %>%
    mutate(
        unique_site_id = base::make.unique(
            str_c(symbol, refseq_prot_id, phosphosites, sep = ':'), 
            sep='.'
        )
    ) %>%
    select(unique_site_id, refseq_prot_id, symbol, phosphosites, 
           peptide, peptide_start, peptide_end, 
           original_id)

phospho_row_full_tbl %>% head()
```

Make the summarized experiment object
```{r}
phospho_mat.selected = phospho_mat[phospho_row_full_tbl$original_id, ]
rownames(phospho_mat.selected) = phospho_row_full_tbl$unique_site_id

phospho_se = SummarizedExperiment(
    assays = list('normalized' = phospho_mat.selected),
    rowData = phospho_row_full_tbl %>% column_to_rownames('unique_site_id'),
    colData = protein_sample_tbl %>% column_to_rownames('preferred_sample_name'),
    metadata = se_metadata
)

phospho_se
```


## Export
```{r}
protein_se %>% saveRDS(here('final_output/CPTAC2_CRC_prospective/CPTAC2_CRC_prospective_pnnl_proteome.rds'))
phospho_se %>% saveRDS(here('final_output/CPTAC2_CRC_prospective/CPTAC2_CRC_prospective_pnnl_phosphoproteome.rds'))
```