---
title: "R Notebook"
output:
    html_document:
        df_print: paged
---

```{r, message=FALSE}
library(SummarizedExperiment)
library(here)
library(readxl)
library(tidyverse)
```

```{r}
DATA_ROOT = '~/Box/MyCPTAC/CPTAC_proteome_v2.0'
```

Load extra functions for mapping
```{r}
source('utils.R')
```

```{r}
tmt_pth = fs::path(DATA_ROOT, 'CPTAC3/HNSCC_discovery/S054_CPTAC_HNSCC_TMT11_Label_to_Sample_Mapping_File_JHU_May2020_r1.xlsx')
protein_pth = fs::path(DATA_ROOT, 'CPTAC3/HNSCC_discovery/UMich-global-proteomics-abundance-MD.tsv.gz')
phospho_pth = fs::path(DATA_ROOT, 'CPTAC3/HNSCC_discovery/Site-level-abundance-MD.tsv.gz')
phospho_loc_pth = fs::path(DATA_ROOT, 'CPTAC3/HNSCC_discovery/Site-level-abundance-MD.peptide_location.tsv.gz')

se_metadata = list(
    cohort = 'CPTAC3 HNSCC discovery cohort',
    pipeline = 'UMich MD',
    mapped_uniprot_release = '2020_04'
)
```

Read UniProt coord_mapping
```{r}
uniprot_coord_mapping_tbl = read_refseq_to_uniprot_coord_mapping(
    refseq_ver = "20180629", 
    uniprot_ver = se_metadata$mapped_uniprot_release
)
```

## TMT map
```{r}
tmt_map_tbl = read_excel(tmt_pth) %>%
    filter(PCC == 'JHU', str_detect(`Folder Name`, 'Proteome'))
tmt_map_tbl %>% head()
```

```{r}
tmt_map_clean_tbl = tmt_map_tbl %>%
    select(folder = `Folder Name`, starts_with('TMT11')) %>%
    pivot_longer(-folder, names_to = c('tmt_channel', 'id_type'), names_pattern = '^TMT11-(\\d+[NC]?) (.*)$') %>%
    pivot_wider(names_from = 'id_type', values_from = 'value') %>%
    filter(tmt_channel != '126') %>%
    mutate_at(vars(`Participant ID`, `Specimen Label`), ~ str_replace(.x, '_disqualified$', '')) %>%
    mutate(tmt_experiment = as.integer(str_match(folder, '^(\\d+)CPTAC_')[, 2]))

tmt_map_clean_tbl
```


## Protein
```{r}
protein_tbl = read_tsv(protein_pth)
```

```{r}
protein_annotation_gr = read_rds(here(
    'annotation/refseq_20180629/tracked_results/refseq_20180629_human_only_unique_loci_GRanges.rds'
))
names(mcols(protein_annotation_gr))
```

```{r}
gene_annotation_tbl = protein_annotation_gr %>%
    as_tibble() %>%
    distinct(symbol, hgnc_id, entrez_gene_id, ensembl_gene_id)
```


Make sure all the genes are unique
```{r}
protein_tbl %>%
    count(Index) %>%
    filter(n > 1)
```
All the genes are in the annotation
```{r}
setdiff(protein_tbl$Index, gene_annotation_tbl$symbol)
```

```{r}
nrow(protein_tbl)
```

```{r}
preferred_sample_names = colnames(protein_tbl)[-(1:2)]
length(preferred_sample_names) == length(tmt_map_clean_tbl$`Participant ID`)
```

```{r}
protein_mat = protein_tbl %>% 
    select(-Index, -ReferenceIntensity) %>%
    as.matrix()

protein_mat = protein_mat - protein_tbl$ReferenceIntensity
rownames(protein_mat) = protein_tbl$Index

protein_mat[1:5, 1:5]
```

```{r}
protein_sample_tbl = bind_cols(
    tibble(
        preferred_sample_names = preferred_sample_names
    ),
    tmt_map_clean_tbl %>% select(
        case_id = `Participant ID`, aliquot_id = `Specimen Label`, 
        tmt_experiment, tmt_channel
    )
) %>%
    mutate(
        sample_type = case_when(
            startsWith(case_id, 'C3') & str_detect(preferred_sample_names, '-T') ~ 'Primary Tumor',
            startsWith(case_id, 'C3') & str_detect(preferred_sample_names, '-N') ~ 'Solid Tissue Normal',
            startsWith(case_id, 'C3') & str_detect(preferred_sample_names, '-C') ~ 'Solid Tissue Normal',
            TRUE ~ NA_character_
        )
    ) %>%
    select(preferred_sample_names, case_id, sample_type, aliquot_id, tmt_experiment, tmt_channel)

protein_sample_tbl %>% head()
```
```{r}
protein_row_tbl = protein_tbl %>%
    select(symbol = Index) %>%
    left_join(gene_annotation_tbl, by = 'symbol')

protein_row_tbl %>% head()
```
```{r}
protein_se = SummarizedExperiment(
    assays = list('normalized' = protein_mat),
    rowData = protein_row_tbl %>% column_to_rownames('symbol'),
    colData = protein_sample_tbl %>% column_to_rownames('preferred_sample_names'),
    metadata = se_metadata
)

protein_se
```



### Phospho
```{r}
phospho_tbl = read_tsv(phospho_pth)
```

```{r}
phospho_tbl %>%
    count(X206)
```
Read peptide location
```{r}
phospho_loc_tbl = read_tsv(
    phospho_loc_pth,
    col_types = cols(
        peptide_start = col_integer(), 
        peptide_end = col_integer()
    ))
phospho_loc_tbl %>% head()
```

```{r}
phospho_loc_tbl %>% 
    count(peptide_loc_valid)
```

Create the phospho matrix
```{r}
phospho_mat = phospho_tbl %>%
    select(-(Index:ReferenceIntensity), -`X206`) %>%
    as.matrix()

phospho_mat = phospho_mat - phospho_tbl$ReferenceIntensity
colnames(phospho_mat) = preferred_sample_names
rownames(phospho_mat) = phospho_tbl$Index

phospho_mat[1:5, 1:5]
```

```{r}
phospho_na_percent = rowSums(is.na(phospho_mat)) / ncol(phospho_mat)
hist(phospho_na_percent)
```


Make sure peptides only consist STY phosphosites
```{r}
flatten_chr(
    str_extract_all(phospho_tbl$Peptide, '[a-z]')
) %>%
    table()
```

Note the peptide start end are not really the start end
```{r}
phospho_row_tbl = phospho_tbl %>%
    select(Index, Gene, Peptide, MaxPepProb) %>%
    separate(
        Index, 
        into = c("refseq_prot_type", "refseq_prot_num", 
                 "peptide_num1", "peptide_num2", 
                 "peptide_num3", "peptide_num4", "phosphosites"),
        sep = "_",
        fill = "right",
        remove = FALSE
    ) %>%
    unite("refseq_prot_id", c("refseq_prot_type", "refseq_prot_num")) %>%
    select(
        original_id = Index,
        refseq_prot_id,
        phosphosites,
        symbol = Gene,
        peptide = Peptide
    ) %>%
    left_join(phospho_loc_tbl, by = c('original_id' = 'Index')) %>%
    bind_cols(tibble(na_percent = phospho_na_percent)) %>%
    # Only keep the phosphosites
    filter(
        !is.na(phosphosites), 
        # na_percent <= 0.8,
        peptide_loc_valid
    ) %>%
    mutate(
        phosphosites_list = str_extract_all(phosphosites, '([STY]\\d+)'),
        phosphosites = map_chr(phosphosites_list, str_c, collapse = ';')
    ) %>%
    # Keep the less NA% peptide as the default site
    arrange(symbol, phosphosites, na_percent) %>%
    mutate(
        unique_site_id = base::make.unique(
            str_c(symbol, refseq_prot_id, phosphosites, sep = ':'), 
            sep='.'
        )
    )
phospho_row_tbl %>% head()
```
Some sites have duplication
```{r}
phospho_row_tbl %>% 
    count(symbol, phosphosites, sort = TRUE) %>% 
    filter(n > 1)
```

Make sure all the RefSeq IDs are available
```{r}
setdiff(
    phospho_row_tbl$refseq_prot_id,
    protein_annotation_gr$refseq_prot_id
)
```

Combine the RefSeq annotation
```{r}
protein_annotation_tbl = protein_annotation_gr %>%
    mcols() %>%
    as_tibble() %>%
    select(-symbol)
```

```{r}
phospho_row_full_tbl = phospho_row_tbl %>% 
    select(unique_site_id, original_id, refseq_prot_id, symbol, phosphosites, 
           peptide, peptide_start, peptide_end) %>%
    left_join(
        protein_annotation_tbl,
        by = 'refseq_prot_id'
    ) %>%
    annotate_uniprot_mapping("phosphosites", uniprot_coord_mapping_tbl)

phospho_row_full_tbl %>% head()
```

Make the summarized experiment object
```{r}
phospho_mat.selected = phospho_mat[phospho_row_full_tbl$original_id, ]
rownames(phospho_mat.selected) = phospho_row_full_tbl$unique_site_id

phospho_se = SummarizedExperiment(
    assays = list('normalized' = phospho_mat.selected),
    rowData = phospho_row_full_tbl %>% column_to_rownames('unique_site_id'),
    colData = protein_sample_tbl %>% column_to_rownames('preferred_sample_names'),
    metadata = se_metadata
)

phospho_se
```

```{r}
samples_to_keep = colData(protein_se) %>%
    as_tibble(rownames = 'preferred_sample_name') %>%
    filter(
        !is.na(sample_type),
        !str_detect(sample_type, '-C')
    ) %>%
    pull(preferred_sample_name)

length(samples_to_keep)
```

```{r}
selected_protein_se = protein_se[, samples_to_keep]
selected_phospho_se = phospho_se[, samples_to_keep]
```



## Export
```{r}
fs::dir_create(here('final_output/CPTAC3_HNSCC_discovery'))
selected_protein_se %>% saveRDS(here('final_output/CPTAC3_HNSCC_discovery/CPTAC3_HNSCC_discovery_umich_proteome.v2.1.rds'))
selected_phospho_se %>% saveRDS(here('final_output/CPTAC3_HNSCC_discovery/CPTAC3_HNSCC_discovery_umich_phosphoproteome.v2.2.rds'))
```

