---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---

```{r, message=FALSE}
library(SummarizedExperiment)
library(here)
library(readxl)
library(tidyverse)
```

```{r}
DATA_ROOT = '~/Box/MyCPTAC/CPTAC_proteome_v2.0'
```

Load extra functions for mapping
```{r}
source('utils.R')
```

```{r}
# tmt_pth = fs::path(DATA_ROOT, 'CPTAC3/UCEC_discovery/S043_CPTAC_UCEC_Discovery_Cohort_Study_Specimens_r1_Sept2018.xlsx')
tmt_pth = fs::path(DATA_ROOT, 'CPTAC3/UCEC_discovery/UCEC_CPTAC3_meta_table_V2.1.txt.gz')

protein_pth = fs::path(DATA_ROOT, 'CPTAC3/UCEC_discovery/PNNL_with_refseq/ucec_global_gene_refseq_B12345_medpolish.txt.gz')
phospho_pth = fs::path(DATA_ROOT, 'CPTAC3/UCEC_discovery/PNNL_with_refseq/ucec_phospho_d6_refseq_B12345_with_sites.txt.gz')
phospho_loc_pth = fs::path(DATA_ROOT, 'CPTAC3/UCEC_discovery/PNNL_with_refseq/ucec_phospho_d6_refseq_B12345_with_sites.txt.peptide_location.tsv.gz')
acetyl_pth = fs::path(DATA_ROOT, 'CPTAC3/UCEC_discovery/PNNL_with_refseq/ucec_acetyl_d6_refseq_B12345_with_sites.txt.gz')
acetyl_loc_pth = fs::path(DATA_ROOT, 'CPTAC3/UCEC_discovery/PNNL_with_refseq/ucec_acetyl_d6_refseq_B12345_with_sites.txt.peptide_location.tsv.gz')
```

Read UniProt 2020_03 coord_mapping
```{r}
uniprot2020_03_coord_mapping_tbl = read_refseq20180629_to_uniprot2020_03_coord_mapping()
```


## TMT map
```{r}
tmt_map_tbl = read_tsv(tmt_pth) %>%
    select(preferred_sample_name = idx, 
           case_id = Proteomics_Participant_ID, 
           aliquot_ids = Proteomics_Aliquot_ID,
           sample_type = Proteomics_Tumor_Normal, 
           OCT = Proteomics_OCT,
           Case_excluded, 
           tmt_experiment = Proteomics_TMT_plex, 
           tmt_channel = Proteomics_TMT_channel) %>%
    mutate(
        tmt_experiment = as.integer(tmt_experiment),
        aliquot_ids = str_split(aliquot_ids, ','),
        first_aliquot_id = map_chr(aliquot_ids, ~ .x[[1]]),
        aliquot_ids = map_chr(aliquot_ids, str_c, collapse = ';')
    )

tmt_map_tbl %>% head()
```

```{r}
tmt_map_clean_tbl = tmt_map_tbl %>%
    filter(Case_excluded == 'No')

tmt_map_clean_tbl
```



## Protein
```{r}
protein_tbl = read_tsv(protein_pth) %>%
    rename_at(vars(-Gene, -RefSeq), ~ str_replace(.x, fixed('.'), ''))
```

```{r}
protein_annotation_gr = read_rds(here(
    'annotation/refseq_20180629/tracked_results/refseq_20180629_human_only_unique_loci_GRanges.rds'
))
names(mcols(protein_annotation_gr))
```

```{r}
gene_annotation_tbl = protein_annotation_gr %>%
    as_tibble() %>%
    distinct(symbol, hgnc_id, entrez_gene_id, ensembl_gene_id)
```


Make sure all the genes are unique
```{r}
protein_tbl %>%
    count(Gene) %>%
    filter(n > 1)
```

All the genes are in the annotation
```{r}
setdiff(protein_tbl$Gene, gene_annotation_tbl$symbol)
```

UCEC is a bit different because the TMT map only includes the samples that are kept. Make sure the full TMT map can be mapped to the samples in data:
```{r}
data_sample_names = colnames(protein_tbl)[-(1:2)]
setdiff(tmt_map_clean_tbl$first_aliquot_id, data_sample_names)
```

```{r}
protein_mat = protein_tbl %>% 
    select(-Gene, -RefSeq) %>%
    as.matrix()

# Re-order the samples
protein_mat = protein_mat[, tmt_map_clean_tbl$first_aliquot_id]
colnames(protein_mat) = tmt_map_clean_tbl$preferred_sample_name
rownames(protein_mat) = protein_tbl$Gene

protein_mat[1:5, 1:5]
```

```{r}
protein_sample_tbl = tmt_map_clean_tbl %>% 
  select(
        preferred_sample_name, case_id, sample_type,
        aliquot_ids, tmt_experiment, tmt_channel, OCT
    )

protein_sample_tbl %>% head()
```

```{r}
protein_row_tbl = protein_tbl %>%
    select(symbol = Gene) %>%
    left_join(gene_annotation_tbl, by = 'symbol')

protein_row_tbl %>% head()
```

```{r}
protein_se = SummarizedExperiment(
    assays = list('normalized' = protein_mat),
    rowData = protein_row_tbl %>% column_to_rownames('symbol'),
    colData = protein_sample_tbl %>% column_to_rownames('preferred_sample_name'),
    metadata = list(
        cohort = 'CPTAC3 UCEC discovery cohort',
        pipeline = 'PNNL'
    )
)

protein_se
```



### Phospho
```{r}
phospho_tbl = read_tsv(phospho_pth) %>%
    rename_at(vars(-Gene:-site), ~ str_replace(.x, fixed('.'), ''))
```

Read peptide location
```{r}
phospho_loc_tbl = read_tsv(
    phospho_loc_pth,
    col_types = cols(
        peptide_start = col_integer(), 
        peptide_end = col_integer()
    ))
phospho_loc_tbl %>% head()
```

Create the phospho matrix
```{r}
phospho_mat = phospho_tbl %>%
    select(-(Gene:site)) %>%
    as.matrix()

# Re-order the samples
phospho_mat = phospho_mat[, tmt_map_clean_tbl$first_aliquot_id]
colnames(phospho_mat) = tmt_map_clean_tbl$preferred_sample_name
rownames(phospho_mat) = phospho_tbl$peptide

phospho_mat[1:5, 1:5]
```

Construct the phospho rowData table
```{r}
raw_phospho_row_tbl = phospho_tbl %>%
    select(original_id = peptide, symbol = Gene, refseq_prot_id = RefSeq) %>% 
    inner_join(phospho_loc_tbl, by = 'original_id')

raw_phospho_row_tbl %>% head()
```

Calculate the NA% per peptide (to order the site)
```{r}
phospho_na_percent = rowSums(is.na(phospho_mat)) / ncol(phospho_mat)
hist(phospho_na_percent)
```

```{r}
phospho_row_tbl = raw_phospho_row_tbl %>%
    bind_cols(tibble(na_percent = phospho_na_percent)) %>%
    filter(
        peptide_loc_valid,
        # na_percent <= 0.8
    ) %>%
    select(
        original_id, 
        refseq_prot_id,
        phosphosites,
        symbol,
        peptide,
        peptide_start,
        peptide_end,
        na_percent
    ) %>%
    arrange(symbol, phosphosites, na_percent) %>%
    mutate(
        unique_site_id = base::make.unique(
            str_c(symbol, refseq_prot_id, phosphosites, sep = ':'), 
            sep='.'
        )
    )

phospho_row_tbl %>% head()
```

Some sites have duplication
```{r}
phospho_row_tbl %>% 
    count(symbol, phosphosites, sort = TRUE) %>% 
    filter(n > 1)
```

Make sure all the RefSeq IDs are available
```{r}
setdiff(
    phospho_row_tbl$refseq_prot_id,
    protein_annotation_gr$refseq_prot_id
)
```

Combine the RefSeq annotation
```{r}
protein_annotation_tbl = protein_annotation_gr %>%
    mcols() %>%
    as_tibble() %>%
    select(-symbol)
```

```{r}
phospho_row_full_tbl = phospho_row_tbl %>% 
    select(unique_site_id, original_id, refseq_prot_id, symbol, phosphosites, 
           peptide, peptide_start, peptide_end) %>%
    left_join(
        protein_annotation_tbl,
        by = 'refseq_prot_id'
    )  %>%
    annotate_uniprot_mapping("phosphosites", uniprot2020_03_coord_mapping_tbl)

phospho_row_full_tbl %>% head()
```

Make the summarized experiment object
```{r}
phospho_mat.selected = phospho_mat[phospho_row_full_tbl$original_id, ]
rownames(phospho_mat.selected) = phospho_row_full_tbl$unique_site_id

phospho_se = SummarizedExperiment(
    assays = list('normalized' = phospho_mat.selected),
    rowData = phospho_row_full_tbl %>% column_to_rownames('unique_site_id'),
    colData = protein_sample_tbl %>% column_to_rownames('preferred_sample_name'),
    metadata = list(
        cohort = 'CPTAC3 UCEC discovery cohort',
        pipeline = 'PNNL'
    )
)

phospho_se
```


## Acetylome
```{r}
acetyl_tbl = read_tsv(acetyl_pth) %>%
    rename_at(vars(-Gene:-site), ~ str_replace(.x, fixed('.'), ''))

acetyl_loc_tbl = read_tsv(
    acetyl_loc_pth,
    col_types = cols(
        peptide_start = col_integer(), 
        peptide_end = col_integer()
    ))
```

Create the phospho matrix
```{r}
acetyl_mat = acetyl_tbl %>%
    select(-(Gene:site)) %>%
    as.matrix()

# Re-order the samples
acetyl_mat = acetyl_mat[, tmt_map_clean_tbl$first_aliquot_id]
colnames(acetyl_mat) = tmt_map_clean_tbl$preferred_sample_name
rownames(acetyl_mat) = acetyl_tbl$peptide

acetyl_mat[1:5, 1:5]
```

Construct the phospho rowData table
```{r}
raw_acetyl_row_tbl = acetyl_tbl %>%
    select(original_id = peptide, symbol = Gene, refseq_prot_id = RefSeq) %>% 
    inner_join(acetyl_loc_tbl, by = 'original_id')

raw_acetyl_row_tbl %>% head()
```

```{r}
acetyl_na_percent = rowSums(is.na(acetyl_mat[raw_acetyl_row_tbl$original_id, ])) / ncol(acetyl_mat)
hist(acetyl_na_percent)
```

```{r}
acetyl_row_tbl = raw_acetyl_row_tbl %>%
    bind_cols(tibble(na_percent = acetyl_na_percent)) %>%
    filter(
        peptide_loc_valid,
        # na_percent <= 0.8
    ) %>%
    select(
        original_id, 
        refseq_prot_id,
        acetylsites,
        symbol,
        peptide,
        peptide_start,
        peptide_end,
        na_percent
    ) %>%
    arrange(symbol, acetylsites, na_percent) %>%
    mutate(
        unique_site_id = base::make.unique(
            str_c(symbol, refseq_prot_id, acetylsites, sep = ':'), 
            sep='.'
        )
    )

acetyl_row_tbl %>% head()
```

Some sites have duplication
```{r}
acetyl_row_tbl %>% 
    count(symbol, acetylsites, sort = TRUE) %>% 
    filter(n > 1)
```


Combine the RefSeq annotation
```{r}
acetyl_row_full_tbl = acetyl_row_tbl %>% 
    select(unique_site_id, original_id, refseq_prot_id, symbol, acetylsites, 
           peptide, peptide_start, peptide_end) %>%
    left_join(
        protein_annotation_tbl,
        by = 'refseq_prot_id'
    )  %>%
    annotate_uniprot_mapping("acetylsites", uniprot2020_03_coord_mapping_tbl)

acetyl_row_full_tbl %>% head()
```
Check if we can re-use the protein col data
```{r}
identical(colnames(acetyl_mat), tmt_map_clean_tbl$preferred_sample_name)
```

```{r}
acetyl_mat.selected = acetyl_mat[acetyl_row_full_tbl$original_id, ]
rownames(acetyl_mat.selected) = acetyl_row_full_tbl$unique_site_id

acteyl_se = SummarizedExperiment(
    assays = list('normalized' = acetyl_mat.selected),
    rowData = acetyl_row_full_tbl %>% column_to_rownames('unique_site_id'),
    colData = protein_sample_tbl %>% column_to_rownames('preferred_sample_name'),
    metadata = list(
        cohort = 'CPTAC3 UCEC discovery cohort',
        pipeline = 'PNNL'
    )
)

acteyl_se
```



## Export
```{r}
fs::dir_create(here('final_output/CPTAC3_UCEC_discovery'))

protein_se %>% saveRDS(here('final_output/CPTAC3_UCEC_discovery/CPTAC3_UCEC_discovery_pnnl_proteome.v2.0.rds'))
phospho_se %>% saveRDS(here('final_output/CPTAC3_UCEC_discovery/CPTAC3_UCEC_discovery_pnnl_phosphoproteome.v2.2.rds'))
acteyl_se %>% saveRDS(here('final_output/CPTAC3_UCEC_discovery/CPTAC3_UCEC_discovery_pnnl_acetylome.v2.2.rds'))
```
