---
title: "R Notebook"
output:
    html_document:
        df_print: paged
---

```{r, message=FALSE}
library(SummarizedExperiment)
library(here)
library(tidyverse)
```

```{r}
DATA_ROOT = '~/Box/MyCPTAC/CPTAC_proteome_v3.0'
```

Load extra functions for mapping
```{r}
source('utils.R')
```

```{r}
tmt_pth = fs::path(DATA_ROOT, 'CPTAC3/CCRCC_discovery/cptac-metadata.csv')

protein_pth = fs::path(DATA_ROOT, 'CPTAC3/CCRCC_discovery/6_CPTAC3_CCRCC_Whole_abundance_gene_protNorm=2_CB.tsv.gz')
phospho_pth = fs::path(DATA_ROOT, 'CPTAC3/CCRCC_discovery/6_CPTAC3_CCRCC_Phospho_abundance_phosphosite_protNorm=2_CB.tsv.gz')
phospho_loc_pth = fs::path(DATA_ROOT, 'CPTAC3/CCRCC_discovery/6_CPTAC3_CCRCC_Phospho_abundance_phosphosite_protNorm=2_CB.tsv.peptide_location.tsv.gz')

se_metadata = list(
    cohort = 'CPTAC3 CCRCC discovery cohort',
    pipeline = 'UMich',
    mapped_uniprot_release = '2021_02'
)
```

Read UniProt coord_mapping
```{r}
uniprot_coord_mapping_tbl = read_refseq_to_uniprot_coord_mapping(
    refseq_ver = "20180629",
    uniprot_ver = se_metadata$mapped_uniprot_release
)
```



## TMT map
```{r}
tmt_map_tbl = read_csv(tmt_pth, col_types = cols(Data.Set = col_integer())) %>%
    select(
        case_id = Case.ID,
        aliquot_id = Specimen.Label,
        sample_type = Type,
        tmt_experiment = Data.Set,
        tmt_channel = TMT.Channel
    ) %>%
    mutate(
        preferred_sample_name = aliquot_id,
        sample_type = case_when(
            sample_type == 'Tumor' ~ 'Primary Tumor',
            sample_type == 'Normal' ~ 'Solid Tissue Normal',
            TRUE ~ NA_character_
        )
    )

tmt_map_tbl %>% head()
```
```{r}
tmt_map_clean_tbl = tmt_map_tbl %>%
    # Remove pool and cell line samples
    filter(!is.na(sample_type))

tmt_map_clean_tbl %>% head()
```


## Protein
```{r}
protein_tbl = read_tsv(protein_pth)
```

```{r}
protein_annotation_gr = read_rds(here(
    'annotation/refseq_20180629/tracked_results/refseq_20180629_human_only_unique_loci_GRanges.rds'
))
names(mcols(protein_annotation_gr))
```

```{r}
gene_annotation_tbl = protein_annotation_gr %>%
    as_tibble() %>%
    distinct(symbol, hgnc_id, entrez_gene_id, ensembl_gene_id)
```


Make sure all the genes are unique
```{r}
protein_tbl %>%
    count(Index) %>%
    filter(n > 1)
```

All the genes are in the annotation
```{r}
setdiff(protein_tbl$Index, gene_annotation_tbl$symbol)
```

```{r}
protein_mat = protein_tbl %>%
    select(-Index, -NumberPSM, -Proteins) %>%
    as.matrix()

# Re-order the samples
protein_mat = protein_mat[, tmt_map_clean_tbl$preferred_sample_name]
protein_mat = protein_mat - protein_tbl$ReferenceIntensity
rownames(protein_mat) = protein_tbl$Index

protein_mat[1:5, 1:5]
```

```{r}
protein_sample_tbl = tmt_map_clean_tbl %>%
  select(
        preferred_sample_name, case_id, sample_type,
        aliquot_id, tmt_experiment, tmt_channel
    )

protein_sample_tbl %>% head()
```

```{r}
protein_row_tbl = protein_tbl %>%
    select(symbol = Index, num_psm = NumberPSM, refseq_prot_ids = Proteins) %>%
    mutate(refseq_prot_ids = map_chr(refseq_prot_ids, str_remove, ';$')) %>%
    left_join(gene_annotation_tbl, by = 'symbol')

protein_row_tbl %>% head()
```

```{r}
protein_se = SummarizedExperiment(
    assays = list('normalized' = protein_mat),
    rowData = protein_row_tbl %>% column_to_rownames('symbol'),
    colData = protein_sample_tbl %>% column_to_rownames('preferred_sample_name'),
    metadata = se_metadata
)

protein_se
```



### Phospho
```{r}
phospho_tbl = read_tsv(
    phospho_pth,
    col_types = cols(
        .default = col_double(),
        Index = col_character(),
        Gene = col_character(),
        Peptide = col_character()
    )
)
```

Read peptide location
```{r}
phospho_loc_tbl = read_tsv(
    phospho_loc_pth,
    col_types = cols(
        peptide_start = col_integer(),
        peptide_end = col_integer()
    ))
phospho_loc_tbl %>% head()
```

Create the phospho matrix
```{r}
phospho_mat = phospho_tbl %>%
    select(-(Index:ReferenceIntensity)) %>%
    as.matrix()

# Re-order the samples
phospho_mat = phospho_mat[, protein_sample_tbl$preferred_sample_name]
phospho_mat = phospho_mat - phospho_tbl$ReferenceIntensity
rownames(phospho_mat) = phospho_tbl$Index

phospho_mat[1:5, 1:5]
```

Construct the phospho rowData table
```{r}
raw_phospho_row_tbl = phospho_tbl %>%
    select(original_id = Index, symbol = Gene) %>%
    left_join(phospho_loc_tbl, by = 'original_id')

raw_phospho_row_tbl %>% head()
```
Some phosphosites are invalid or they are not phosphosites at all
```{r}
raw_phospho_row_tbl %>% count(peptide_loc_valid)
```


Calculate the NA% per peptide (to order the site)
```{r}
phospho_na_percent = rowSums(is.na(phospho_mat)) / ncol(phospho_mat)
hist(phospho_na_percent)
```

```{r}
phospho_row_tbl = raw_phospho_row_tbl %>%
    bind_cols(tibble(na_percent = phospho_na_percent)) %>%
    filter(
        peptide_loc_valid,
        # na_percent <= 0.8
    ) %>%
    select(
        original_id,
        refseq_prot_id,
        phosphosites,
        symbol,
        peptide,
        peptide_start,
        peptide_end,
        na_percent
    ) %>%
    arrange(symbol, phosphosites, na_percent) %>%
    mutate(
        unique_site_id = base::make.unique(
            str_c(symbol, refseq_prot_id, phosphosites, sep = ':'),
            sep='.'
        )
    )

phospho_row_tbl %>% head()
```

Some sites have duplication
```{r}
phospho_row_tbl %>%
    count(symbol, phosphosites, sort = TRUE) %>%
    filter(n > 1)
```

Make sure all the RefSeq IDs are available
```{r}
setdiff(
    phospho_row_tbl$refseq_prot_id,
    protein_annotation_gr$refseq_prot_id
)
```

Combine the RefSeq annotation
```{r}
protein_annotation_tbl = protein_annotation_gr %>%
    mcols() %>%
    as_tibble() %>%
    select(-symbol)
```

```{r}
phospho_row_full_tbl = phospho_row_tbl %>%
    select(unique_site_id, original_id, refseq_prot_id, symbol, phosphosites,
           peptide, peptide_start, peptide_end) %>%
    left_join(
        protein_annotation_tbl,
        by = 'refseq_prot_id'
    ) %>%
    annotate_uniprot_mapping("phosphosites", uniprot_coord_mapping_tbl)

phospho_row_full_tbl %>% head()
```

Make the summarized experiment object
```{r}
phospho_mat.selected = phospho_mat[phospho_row_full_tbl$original_id, ]
rownames(phospho_mat.selected) = phospho_row_full_tbl$unique_site_id

phospho_se = SummarizedExperiment(
    assays = list('normalized' = phospho_mat.selected),
    rowData = phospho_row_full_tbl %>% column_to_rownames('unique_site_id'),
    colData = protein_sample_tbl %>% column_to_rownames('preferred_sample_name'),
    metadata = se_metadata
)

phospho_se
```



## Export
```{r}
fs::dir_create(here('final_output/CPTAC3_CCRCC_discovery'))

protein_se %>% saveRDS(here('final_output/CPTAC3_CCRCC_discovery/CPTAC3_CCRCC_discovery_umich_proteome.v2.0.rds'))
phospho_se %>% saveRDS(here('final_output/CPTAC3_CCRCC_discovery/CPTAC3_CCRCC_discovery_umich_phosphoproteome.v2.2.rds'))
```
